How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-27 docs/PhysicsPaper/StokesErosion.tex
NM

% New
\newcommand{\anis}{\mathcal{A}}
\newcommand{\diag}{\mathop{\mathrm{diag}}}

@@

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anis = T_2/T_1.
\end{align}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig2.pdf}
\caption{A single simulation of porous-medium erosion. The initial configuration consists of 80 circular bodies of random size and position, as seen in Fig.~\ref{fig1}. a) As the medium erodes, the fraction of solid-body area decreases, or, equivalently, the porosity increases. b) Resistance to flow can be characterized by either the inverse permeability (solid) or the cumulative drag (dots), the two of which are related through \eqref{perm_drag}. Both are seen to decrease as the medium erodes, and relationship \eqref{perm_drag} is confirmed by the simulation data. The figure also shows the inverse permeability in transverse direction (dashed).
 \label{fig2}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
\caption{
Permeability
\label{fig3}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4
%c) The relationship between permeability $k$ and porosity $\phi$ for the eroding medium. Both quantities increase with time as the medium erodes, but at different rates due to changes in the microstructure and channelization of the medium.

@@

>>>>

% New
\newcommand{\anis}{\mathcal{A}}
\newcommand{\anistort}{\mathcal{A_T}}
\newcommand{\diag}{\mathop{\mathrm{diag}}}

@@

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
\todo[inline]{What is $x_0$ in the denominator? Should that be -1?}
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees and compute a transverse tortuosity $T_2$. Then, another measure of the anisotropy is
\begin{align}
  \anistort = T_2/T_1.
\end{align}
As in the permeability definition of anisotropy, the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one. As this configuration erodes, the streamlines in the longitudinal direction are expected to less tortuous than those in the transverse direction, yielding $\anis > 1$.

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig2.pdf}
\caption{A single simulation of porous-medium erosion. The initial configuration consists of 80 circular bodies of random size and position, as seen in Fig.~\ref{fig1}. a) As the medium erodes, the fraction of solid-body area decreases, or, equivalently, the porosity increases. b) Resistance to flow can be characterized by either the resistivity (solid) or the cumulative drag (dots), the two of which are related through \eqref{perm_drag}. Both are seen to decrease as the medium erodes, and relationship \eqref{perm_drag} is confirmed by the simulation data. The figure also shows the resistivity in transverse direction (dashed).
 \label{fig2}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

@@

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
\caption{
The permeability and anisotropy of the same 80-body simulation featured in Figs.~\ref{fig1} and \ref{fig2}. a) Both the longitudinal (solid) and transverse (dashed) permeability increase over time as the bodies erode. Also shown in the figure is the longitudinal and transverse permeabilities of the corresponding configurations of circles, which allows us to extract the configurational component of anisotropy. b) The anisotropy of the medium increases as the bodies erode and preferentially allow flow in the longitudinal direction. The total anisotropy is the sum of configurational and shape components. The shape contribution is larger throughout most of the simulation, but near the end of the simulation, the configurational contribution grows large as the channel-structure dominates.
\label{fig3}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4
%c) The relationship between permeability $k$ and porosity $\phi$ for the eroding medium. Both quantities increase with time as the medium erodes, but at different rates due to changes in the microstructure and channelization of the medium.

@@

% Could reference Eastham paper on including chemistry, precipitation reaction.
% Could discuss convection in conjunction with erosion.
